---
title: "Formula 1 2024 <br> Spanish Grand Prix"
subtitle: "Results and performance analysis"
author: "Francesco Zigotti"
date: "12-09-24"
output: 
  ioslides_presentation:
    css: slides.css
    widescreen: yes
    incremental: yes

editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align='center', message=FALSE, warning=FALSE)

library(tidyverse)
library(gganimate)

# Datasets used for the first part of the project
race_results <- read_csv("./Datasets/F1_2024_Results/Formula1_2024season_raceResults.csv")
sprint_results <- read_csv("./Datasets/F1_2024_Results/Formula1_2024season_sprintResults.csv")
quali_results <- read_csv("./Datasets/F1_2024_Results/Formula1_2024season_qualifyingResults.csv")

# Datasets used for the last part of the project
races <- read_csv("./Datasets/F1_2024_Spanish_race_data/races.csv")
lap_times <- read_csv("./Datasets/F1_2024_Spanish_race_data/lap_times.csv")
drivers <- read_csv("./Datasets/F1_2024_Spanish_race_data/drivers.csv")

# Official Formula 1 teams colors from https://www.reddit.com/r/formula1/comments/1avhmjb/f1_2024_hex_codes/
teams_cols <- c("Alpine Renault" = "#FF87BC", "Aston Martin Aramco Mercedes" = "#229971", "Ferrari" = "#E8002D",
          "Haas Ferrari" = "#B6BABD", "Kick Sauber Ferrari" = "#52E252", "McLaren Mercedes" = "#FF8000",
          "Mercedes" = "#27F4D2", "RB Honda RBPT" = "#6692FF", "Red Bull Racing Honda RBPT" = "#3671C6",
          "Williams Mercedes" = "#64C4FF")

drivers_cols <- c("Charles Leclerc" = "#E8002D", "Lando Norris" = "#FF8000", "George Russell" = "#27F4D2", 
          "Max Verstappen" = "#3671C6")

# Removing data from GPs that take place after the Spanish Grand Prix
race_results <- race_results |>
  filter(Track != "Spain" & Track != "Austria" & Track != "Great Britain" & Track != "Hungary" & Track != "Belgium")

sprint_results <- sprint_results |>
  filter(Track != "Austria")

# Let's create a dataframe that tracks the points gained during the season
race_points <- race_results |>
  select(Driver, Team, Points) |>
  rename(Race_points = Points)

sprint_points <- sprint_results |>
  select(Driver, Team, Points) |>
  rename(Sprint_points = Points)

# Let's determine a Grand Prix order to the results (based on event dates)
ordered_race_results <- race_results |>
  mutate(GP_num = case_when(Track == "Bahrain" ~ 1,
                            Track == "Saudi Arabia" ~ 2,
                            Track == "Australia" ~ 3,
                            Track == "Japan" ~ 4,
                            Track == "China" ~ 5,
                            Track == "Miami" ~ 6,
                            Track == "Emilia Romagna" ~ 7,
                            Track == "Monaco" ~ 8,
                            Track == "Canada" ~ 9,
                            Track == "Spain" ~ 10,
                            Track == "Austria" ~ 11,
                            Track == "Great Britain" ~ 12,
                            Track == "Hungary" ~ 13,
                            Track == "Belgium" ~ 14))

ordered_sprint_results <- sprint_results |>
  mutate(GP_num = case_when(Track == "Bahrain" ~ 1,
                            Track == "Saudi Arabia" ~ 2,
                            Track == "Australia" ~ 3,
                            Track == "Japan" ~ 4,
                            Track == "China" ~ 5,
                            Track == "Miami" ~ 6,
                            Track == "Emilia Romagna" ~ 7,
                            Track == "Monaco" ~ 8,
                            Track == "Canada" ~ 9,
                            Track == "Spain" ~ 10,
                            Track == "Austria" ~ 11,
                            Track == "Great Britain" ~ 12,
                            Track == "Hungary" ~ 13,
                            Track == "Belgium" ~ 14))

# Select the data OF the Spain Grand Prix for Qualifying analysis
q3_results <- quali_results |>
  filter(Track == "Spain" & (!is.na(quali_results$Q3)) & quali_results$Q3 != "DNF") |>
  select(Driver, Team, Q3)

# Add drivers name abbreviation for qualifying analysis
q3_results <- q3_results |>
  mutate(Driver = case_when(Driver == "Lando Norris" ~ "NOR", 
                            Driver == "Max Verstappen" ~ "VER",
                            Driver == "Lewis Hamilton" ~ "HAM",
                            Driver == "George Russell" ~ "RUS",
                            Driver == "Charles Leclerc" ~ "LEC",
                            Driver == "Carlos Sainz" ~ "SAI",
                            Driver == "Pierre Gasly" ~ "GAS",
                            Driver == "Sergio Perez" ~ "PER",
                            Driver == "Esteban Ocon" ~ "OCO"))

# Now let's operate some trasformations about the 2nd dataset

# Select the 2024 Spanish Grand Prix from the races list dataset
spanish_grand_prix <- races |>
  filter(year == 2024 & name == "Spanish Grand Prix") |>
  select(raceId)

# Retrieve the raceId corresponding to the 2024 Spanish Grand Prix
target_race_id <- spanish_grand_prix$raceId

# Obtaining the 2024 spanish race lap times from the race lap times dataset
race_lap_times <- lap_times |>
  filter(raceId == target_race_id)

# Let's transform the drivers list in order to mantain the same data format of the results datasets
drivers <- drivers |>
  mutate(driver = paste(forename, surname, sep = " ")) |>
  select(driverId, driver)

# Removing unused columns
race_lap_times <- race_lap_times |>
  select(driverId, lap, milliseconds)

# Join the race lap times with drivers
race_lap_times <- race_lap_times %>%
  inner_join(drivers)

# Adding "team" attribute to the dataframe
race_lap_times <- race_lap_times |>
  select(driver, lap, milliseconds) |>
  mutate(team = case_when(driver == "Max Verstappen" ~ "Red Bull Racing Honda RBPT",
                          driver == "Sergio Pérez" ~ "Red Bull Racing Honda RBPT",
                          driver == "Lewis Hamilton" ~ "Mercedes",
                          driver == "George Russell" ~ "Mercedes",
                          driver == "Charles Leclerc" ~ "Ferrari",
                          driver == "Carlos Sainz" ~ "Ferrari",
                          driver == "Lando Norris" ~ "McLaren Mercedes",
                          driver == "Oscar Piastri" ~ "McLaren Mercedes",
                          driver == "Fernando Alonso" ~ "Aston Martin Aramco Mercedes",
                          driver == "Lance Stroll" ~ "Aston Martin Aramco Mercedes",
                          driver == "Pierre Gasly" ~ "Alpine Renault",
                          driver == "Esteban Ocon" ~ "Alpine Renault",
                          driver == "Yuki Tsunoda" ~ "RB Honda RBPT",
                          driver == "Daniel Ricciardo" ~ "RB Honda RBPT",
                          driver == "Nico Hülkenberg" ~ "Haas Ferrari",
                          driver == "Kevin Magnussen" ~ "Haas Ferrari",
                          driver == "Alexander Albon" ~ "Williams Mercedes",
                          driver == "Logan Sargeant" ~ "Williams Mercedes",
                          driver == "Valtteri Bottas" ~ "Kick Sauber Ferrari",
                          driver == "Guanyu Zhou" ~ "Kick Sauber Ferrari"))

```

## Introduction

This data science project concerns the Formula 1 2024 season, focusing in particular on the **Spanish Grand Prix**. The **Spanish Grand Prix** was held across the weekend of 21-23 June 2024 at the **Circuit de Barcelona-Catalunya** in **Montmeló**, Spain. It was the 10th round of the F1 2024 World Championship.

Historically, this Grand Prix is often an important event for the engineering developments of the cars, exploiting the features of the track that are very helpful to understand aerodynamics upgrades. For this reason, this weekend could be a crucial turning point for the remainder part of the season.

The project consists of two parts: the first part shows the background of the **Spanish Grand Prix**, going into the details of the competition results achieved before this event. The second part analyzes the on-track performances of the Spanish race, highlighting the hierarchies of the contenders.

## Datasets

This project uses two distinct datasets:

1. [Formula 1 2024 - Results](https://github.com/toUpperCase78/formula1-datasets)
2. [Formula 1 2024 - Spanish race data](https://www.kaggle.com/datasets/rohanrao/formula-1-world-championship-1950-2020)

The first one contains the data of the final classifications of the previous Grand Prix. The second one contains the performance data of the **Spanish Grand Prix**, storing the lap times of the drivers during the race session.

# Part 1 <br> Overview of the previous results

## Drivers Standings

Let's show the drivers standings before the **Spanish Grand Prix**

```{r drivers_standings, out.width="80%"}

# Summing up the points based on drivers 
race_drivers_standings <- race_points |>
  group_by(Driver, Team) |>
  summarize(
    Race_points = sum(Race_points),
    .groups = "drop"
  )

sprint_drivers_standings <- sprint_points |>
  group_by(Driver) |>
  summarize(
    Sprint_points = sum(Sprint_points)
  )

# In Formula 1 there are two types of races: classical races and sprint races (in a less number). Sprint races are long 1/3 of a classical race length and they award a lower number of points. So it's needed to compute the sum of the points awarded from both race types.
drivers_standings <- race_drivers_standings %>%
  full_join(sprint_drivers_standings)

# There is a driver that raced in a classical race but not in any sprint race, so it's needed to set at zero the points awarded in sprint races instead of NA
drivers_standings[is.na(drivers_standings)] <- 0

# Summing up the overall points
drivers_standings <-  drivers_standings |>
  mutate(
    Points = Race_points + Sprint_points
  )

# Plotting the Drivers Standings
ggplot(
  drivers_standings,
  aes(x = Points, y = reorder(Driver, Points), fill = Team)
) +
  geom_bar(stat = "Identity", color = "black") +
  geom_text(aes(label = Points), nudge_x = 10) +
  scale_fill_manual(values = teams_cols) +
  labs(
    title = "Drivers standings",
    x = "Points", y = "Driver"
  )
```

## Constructors Standings 

Let's show the constructors standings before the **Spanish Grand Prix**

```{r constructors_standings, out.width="80%"}

# Summing up the points based on drivers 
race_constructors_standings <- race_points |>
  group_by(Team) |>
  summarize(
    Race_points = sum(Race_points)
  )

sprint_constructors_standings <- sprint_points |>
  group_by(Team) |>
  summarize(
    Sprint_points = sum(Sprint_points)
  )

# Joining classic races points and sprint races points
constructors_standings <- race_constructors_standings %>%
  full_join(sprint_constructors_standings)

# Summing up the overall points
constructors_standings <-  constructors_standings |>
  mutate(
    Points = Race_points + Sprint_points
  )

# Plotting the Constructors Standings
ggplot(
  constructors_standings,
  aes(x = Points, y = reorder(Team, Points))
) +
  geom_bar(stat = "Identity", fill = teams_cols, color = "black") +
  geom_text(aes(label = Points), nudge_x = 10) +
  labs(
    title = "Constructors standings",
    x = "Points", y = "Team"
  )
```

## Performance classes

Using these standings, the drivers are divided into performance classes using hierarchical clustering

```{r driver_performance classes, out.width="75%"}

drivers_rank <- drivers_standings |>
  select(Driver, Points)

# Group drivers into performance classes 
drivers_rank = arrange(drivers_rank, desc(Points))
D = dist(drivers_rank$Points)
cc = hclust(D, method = "average")

# Visualize the driver ranking highlighting the performance classes with different colors
clusters = cutree(cc, k = 6)
drivers_rank = mutate(drivers_rank, cluster = as.factor(clusters))
ggplot(drivers_rank, aes(x = 1, y = Points, color = cluster)) +
  geom_text(aes(label = Driver), show.legend = FALSE) + 
  labs(y = "Rating") +
  theme_minimal() + 
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

This computation is repeated also for constructors

```{r constructors_performance classes, out.width="75%"}

constructors_rank <- constructors_standings |>
  select(Team, Points)

# Group teams into performance classes 
constructors_rank = arrange(constructors_rank, desc(Points))
D = dist(constructors_rank$Points)
cc = hclust(D, method = "average")

## Visualize the team ranking highlighting the performance classes with different colors
clusters = cutree(cc, k = 4)
constructors_rank = mutate(constructors_rank, cluster = as.factor(clusters))
ggplot(constructors_rank, aes(x = 1, y = Points, color = cluster)) +
  geom_text(aes(label = Team), show.legend = FALSE) + 
  labs(y = "Rating") +
  theme_minimal() + 
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

## Title contenders drivers Trend Analysis

The previous plots show the overall results in a fixed point (the situation immediately before the **Spanish Grand Prix**). But it's also interesting analyze how they changed during the ongoing season

```{r drivers_title_contenders_trend, out.width="70%"}

# Using the ordered results for this block
ordered_race_results <- ordered_race_results|>
  select(GP_num, Driver, Team, Points)

ordered_sprint_results <- ordered_sprint_results|>
  select(GP_num, Driver, Team, Points)

ordered_race_results <-  rename(ordered_race_results, Race_points = Points)

ordered_sprint_results <-  rename(ordered_sprint_results, Sprint_points = Points)

# Join the classical races and the sprint races results
ordered_results <- ordered_race_results %>%
  full_join(ordered_sprint_results)

ordered_results[is.na(ordered_results)] <- 0

# Summing up the overall points
drivers_points <-  ordered_results |>
  mutate(
    Points = Race_points + Sprint_points
  ) |>
  select(GP_num, Driver, Team, Points)

# In this section we consider only the points of the four "top teams" "number one" drivers (drivers that have more points compared to his teammate)
drivers_title_contenders_points <- drivers_points |>
  filter(Driver == "Max Verstappen" | Driver == "Charles Leclerc" | Driver == "Lando Norris" | Driver == "George Russell")

# Compute the cumulative points of target drivers
cumulative_points_ver <- drivers_title_contenders_points |>
  filter(Driver == "Max Verstappen") |>
  mutate(Cumulative_points = cumsum(Points))

cumulative_points_lec <- drivers_title_contenders_points |>
  filter(Driver == "Charles Leclerc") |>
  mutate(Cumulative_points = cumsum(Points))

cumulative_points_nor <- drivers_title_contenders_points |>
  filter(Driver == "Lando Norris") |>
  mutate(Cumulative_points = cumsum(Points))

cumulative_points_rus <- drivers_title_contenders_points |>
  filter(Driver == "George Russell") |>
  mutate(Cumulative_points = cumsum(Points))

# Concatenate the cumulative points of the four target drivers
cumulative_points <- rbind(cumulative_points_ver, cumulative_points_nor, cumulative_points_lec, cumulative_points_rus)

# Tidying the data on the Grand Prix dates
cumulative_points <- cumulative_points |>
  arrange(GP_num, Driver)

# Visualize the title contenders drivers trend
drivers_trend <- ggplot(
  cumulative_points, 
  aes(x = GP_num, y = Cumulative_points, color = Driver)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 2.5) +
  scale_color_manual(values = drivers_cols) +
  transition_reveal(GP_num)  +
  labs(
    title = "Title contenders drivers trend",
    x = "Number of races", y = "Cumulative points"
  ) + 
  scale_x_continuous(breaks = 1:9)

animate(
  drivers_trend,
  renderer = av_renderer()
)
```

## Title contenders constructors Trend Analysis

Let's have a look also to the constructors results

```{r constructors_title_contenders_trend, out.width="80%"}

# Group the points based on teams
constructors_title_contenders_points <- drivers_points |>
  select(GP_num, Team, Points) |>
  group_by(GP_num, Team) |>
  summarise(Points = sum(Points),
            .groups = "drop"
            )

# In this section we consider just the points of the four "top teams"
constructors_title_contenders_points <- constructors_title_contenders_points |>
  filter(Team == "Red Bull Racing Honda RBPT" | Team == "McLaren Mercedes" | Team == "Ferrari" | Team == "Mercedes")

# Compute the cumulative points of target teams
cumulative_points_rb <- constructors_title_contenders_points |>
  filter(Team == "Red Bull Racing Honda RBPT") |>
  mutate(Cumulative_points = cumsum(Points))

cumulative_points_mcl <- constructors_title_contenders_points |>
  filter(Team == "McLaren Mercedes") |>
  mutate(Cumulative_points = cumsum(Points))

cumulative_points_fer <- constructors_title_contenders_points |>
  filter(Team == "Ferrari") |>
  mutate(Cumulative_points = cumsum(Points))

cumulative_points_mer <- constructors_title_contenders_points |>
  filter(Team == "Mercedes") |>
  mutate(Cumulative_points = cumsum(Points))

# Concatenate the cumulative points of the four target drivers
teams_cumulative_points <- rbind(cumulative_points_rb, cumulative_points_mcl, cumulative_points_fer, cumulative_points_mer)

# Tidying the data on the Grand Prix dates
teams_cumulative_points <- teams_cumulative_points |>
  arrange(GP_num, Team)

# Visualize the title contenders teams trend
teams_trend <- ggplot(
  teams_cumulative_points, 
  aes(x = GP_num, y = Cumulative_points, color = Team)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 2.5) +
  scale_color_manual(values = teams_cols) +
  transition_reveal(GP_num) +
  labs(
    title = "Title contenders teams trend",
    x = "Number of races", y = "Cumulative points"
  ) + 
  scale_x_continuous(breaks = 1:9)

animate(
  teams_trend,
  renderer = av_renderer()
)
```

# Part 2 <br> Spanish Grand Prix performances

## Are qualifying sessions worth to be considered?

There is a correlation between grid position and final race position?

```{r qualifying_relevance, out.width="80%"}

# Retrieve the final race position and the starting grid position 
race_results <- race_results |>
  select(Position, `Starting Grid`)

sprint_results <- sprint_results |>
  select(Position, `Starting Grid`)

# Concatenate the data of classical races and sprint races
results <- rbind(race_results, sprint_results)

# Let's remove the non classified and disqualified drivers
results <- results |>
  filter(Position != "NC" & Position != "DQ")

results$Position <- as.integer(results$Position)

# Visualize a scatterplot to find a correlation between final position and starting position
ggplot(
  results,
  aes(x = Position, y = `Starting Grid`)
) +
  geom_point() + 
  geom_abline() 
```

Looks like there is a positive linear correlation between the starting position and the final position.
To have a confirmation about that, let's compute the **Pearson correlation coefficient** $R$: 

```{r pearson_coefficient}
r_value = cor(results$Position, results$`Starting Grid`, , method = "pearson")
```

* $R$ = `r r_value`

## Qualifying performances

Performances of the last part of the qualifying session at **Spanish Grand Prix**

```{r qualifying_performances, out.width="80%"}

# Converting lap time from Char type to double
q3_results <- q3_results |>
  mutate(Time_ms = (60 + (as.double(str_split_i(Q3, ":", -1)))))

# Visualize the qualifying lap time gap to the pole sitter
ggplot(
  q3_results,
  aes(x = reorder(Driver, Time_ms), y = Time_ms - min(Time_ms), fill = Team)
) + 
  geom_bar(stat = "Identity", color = "black") +
  geom_text(aes(label = round(Time_ms - min(Time_ms), digits = 3)), nudge_y = 0.03) +
  scale_fill_manual(values = teams_cols) +
  labs(
    title = "Qualifying lap time",
    x = "Drivers", y = "Delta to pole position (seconds)"
  )
```

## Race lap times performances

Let's visualize the lap times registered during the whole race by the title contenders drivers

```{r race_lap_times_performances, out.width="75%"}

# Starting from this chunk we will use the 2nd dataset related to spanish race data

# In this section we consider just the points of the four "top teams" "number one" drivers (drivers that have more points compared to his teammate)
title_contenders_spanish_lap_times <- race_lap_times |>
  filter(driver == "Max Verstappen" | driver == "Lando Norris" | driver == "Charles Leclerc" | driver == "George Russell")

# Visualize the line chart of the title contenders drivers lap times
ggplot(
  title_contenders_spanish_lap_times,
  aes(x = lap, y = milliseconds, color = driver)
) + 
  geom_line() +
  scale_color_manual(values = drivers_cols) +
  labs(
    title = "Title contenders drivers lap times",
    x = "Lap number", y = "Lap time"
  )
```

## Race lap times distribution

Let's have a look also to the distribution of the lap times

```{r race_lap_times_distribution, out.width="80%"}

# Visualize the box plot of the title contenders drivers lap times
ggplot(
  title_contenders_spanish_lap_times,
  aes(x = driver, y = milliseconds, color = driver)
) + 
  geom_boxplot(outlier.shape = NA) +
  scale_color_manual(values = drivers_cols) +
  coord_cartesian(ylim = c(77000, 84500)) +
  labs(
    title = "Title contenders drivers lap times distribution",
    x = "Driver", y = "Lap time"
  )
```

## Race gaps to the race winner

The overall driver performance is obtained computing the final gaps respect to the race winner 

```{r driver_overall_race_performace, out.width="75%"}

# Computing the average lap times
average_lap_times <- race_lap_times |>
  group_by(driver, team) |>
  summarise(
    average_lap_time = mean(milliseconds),
    .groups = "drop"
  )

# Visualize the final gaps respect to the race winner
ggplot(
  average_lap_times,
  aes(x = average_lap_time - min(average_lap_time), y = reorder(driver, average_lap_time), fill = team)
) + 
  geom_bar(stat = "Identity", color = "black") +
  scale_fill_manual(values = teams_cols) +
  labs(
    title = "Final gaps",
    x = "Final gap to the race winner", y = "Driver"
  )
```

It's possible to compute the mean between the average lap time of drivers who race for the same team, visualizing the average gaps respect to the fastest average lap times team

```{r team_overall_race_performace, out.width="75%"}

# Computing the mean of average lap times of drivers of the same team
average_team_lap_times <- average_lap_times |>
  group_by(team) |>
  summarise(
    average_lap_time = mean(average_lap_time)
  )

# Visualize the average gaps respect to the fastest average lap times team
ggplot(
  average_team_lap_times,
  aes(x = average_lap_time - min(average_lap_time), y = reorder(team, average_lap_time), fill = team)
) + 
  geom_bar(stat = "Identity", color = "black", show.legend = FALSE) +
  scale_fill_manual(values = teams_cols) +
  labs(
    title = "Average team lap times",
    x = "Average team lap times", y = "Gap to the fastest average team"
  )
```

## Spanish Grand Prix performance classes

In conclusion, let's divide the drivers performances into classes using hierachical clustering

```{r drivers_spanish_performances_classes, out.width="75%"}

# Group drivers into performance classes 
average_lap_times <- arrange(average_lap_times, average_lap_time)
D = dist(average_lap_times$average_lap_time)
cc = hclust(D, method = "average")

# Visualize the driver ranking highlighting the performance classes with different colors
clusters = cutree(cc, k = 6)
average_lap_times = mutate(average_lap_times, cluster = as.factor(clusters))
ggplot(average_lap_times, aes(x = 1, y = desc(average_lap_time), color = cluster)) +
  geom_text(aes(label = driver), show.legend = FALSE) + 
  labs(y = "Rating") +
  theme_minimal() + 
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

The computation is repeated to divide teams into classes

```{r teams_spanish_performances_classes, out.width="75%"}

# Group teams into performance classes 
average_team_lap_times <- arrange(average_team_lap_times, average_lap_time)
D = dist(average_team_lap_times$average_lap_time)
cc = hclust(D, method = "average")

# Visualize the team ranking highlighting the performance classes with different colors
clusters = cutree(cc, k = 4)
average_team_lap_times = mutate(average_team_lap_times, cluster = as.factor(clusters))
ggplot(average_team_lap_times, aes(x = 1, y = desc(average_lap_time), color = cluster)) +
  geom_text(aes(label = team), show.legend = FALSE) + 
  labs(y = "Rating") +
  theme_minimal() + 
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

